Quality Criteria for SKOS Vocabularies

### 2012-03-20 criteria.tex
added computation info for the criteria

\subsubsection{Omitted Language Tags}
Some controlled vocabularies contain literals in natural language, but without information what language has actually been used. Using private language tags (see RFC3066) or omitting them at all limits, e.g., language-dependent queries.

\subsubsection{Incomplete Language Coverage}
As stated above, vocabularies should combine plain literals with language tags. Furthermore, to support, e.g., internationalization or translation use cases, every concept should have defined literal values in the same set of languages than the other concepts.

\subsubsection{Undocumented Concepts}
\cite{Aitchison2000} cites Svenonius to advocate ``inclusion of as much definition material as possible''. The SKOS language specification defines various properties that are intended to hold this kind of information, subsumed as ``documentation properties'' in the SKOS reference.

\subsubsection{Potentially Semantically Related Concepts} \todo{All}{reformulate criterion title}
This criterion is an extension of a statement from the SKOS primer, recommending ``no two concepts have the same preferred lexical label in a given language when they belong to the same concept scheme''. Given the fact that (i) many vocabularies don’t make use of concept schemes and (ii) also identical altLabels or hiddenLabels can have a negative effect in some use-cases (e.g, auto-completion of concept labels based on user input), we follow a more general approach: identification of all pairs of concepts with their respective prefLabel, altLabel or hiddenLabel meeting a threshold of a certain similarity function $sim:LV \times LV \rightarrow [0,1]$. For the analysis of the criterion in this document, we define the similarity function  to check for case-insensitive string equality and the threshold to equal 1.

@@

\subsubsection{Loose Concepts} is a criterion motivated by the notion of ``orphan terms'' in the literature~\cite{Hedden2010}, which are are terms without any associative or hierarchical relationships. Checking for them is common in existing thesaurus development applications and also suggested by \cite{Z39.19:2005}. When we apply this to SKOS vocabularies, a loose concept is a concept that has no semantic relations with other conceptual resources and thus do not occur in any of the pairs yielded by $SR$. Although they might have attached lexical labels, they lack valuable context information that is considered essential for, e.g., search query expansion.

@@

\subsubsection{Omitted Top Concepts}
The SKOS language provides ConceptSchemes which are a facility for grouping related concepts. In order to provide entry points to such a group of concepts, one or more concepts can be marked as top concepts. This helps to provide ``efficient access'' (SKOS primer) and simplifies orientation in the vocabulary.

@@

\subsubsection{Top Concept Having Broader Concepts}
\cite{Allemang2011} proposes to ``not indicate any concepts internal to the tree as top concepts'', i. e. top concepts should not have broader concepts.

@@

\subsubsection{Missing In-Links}
A periodically updated version of a diagram visualizing ``the data sets in the LOD cloud as well as their interlinkage relationships'' is publised online\footnote{\url{http://www4.wiwiss.fu-berlin.de/lodcloud/state/}}. In these diagrams, dbpedia is shown as the most intensely linked resource, i.e. a large number of vocabularies link to resources in that dataset. Therefore a dataset with a large number of other datasets referencing it, can be assumed to be of great value for the community. Accordingly, estimating the number of In-Links of SKOS concepts defined in a vocabulary (henceforth called the in-degree of a concept) would give some hint about how established or accepted a controlled vocabulary is. A method to carry out these estimations would be utilizing existing Linked Data indices like, e.g., Sindice\footnote{\url{http://sindice.com/}} that provide a SPARQL endpoint. Accordingly, this criterion identifies the set of all conceptual resources that lack any In-Links.

\subsubsection{Missing Out-Links}
The Web of Data consists of globally interconnected datasets, ``enabling seamless connections between data sets''\cite{Heath2011}. These links can be established between various SKOS vocabularies by Out-Links, i.e., references from conceptual resources in the ``source vocabulary'' $V$ to resources identified by an HTTP URI and available at a different host. Similar to the criterion above, this criterion identifies the set of all conceptual resources that lack such Out-Links.

@@

\subsubsection{Link Target Unavailability}
This criterion identifies a set of all resources that have been identified as unavailable, i.e. resolving their URI leads to an erroneous HTTP response or no response at all. An erroneous HTTP response in that case can be defined as a response code (after a possible redirection) other than 200. Just as in the ``traditional'' Web, these ``broken links'' hinder information gathering and should be avoided.

@@

\subsubsection{Undefined SKOS Property Usage}
Some vocabularies reference resources in the SKOS namespace that cannot be resolved. A common reason for this is, that there seems to exist a misunderstanding about the correct way of introducing new classes or properties in an OWL document. In some cases, vocabularies ``invent'' new terms in the SKOS namespace in order to meet certain requirements of the publishing organizations that cannot (yet) be expressed within the SKOS schema. Another reason for the presence of unresolvable resource references in the SKOS namespace is that the vocabulary might be outdated. With the SKOS W3C recommendation maturing, some properties have been removed from the current version of the schema. Vocabularies using these ``deprecated'' properties hence contain parts of their information in a non ``standard"-compliant way, having negative effects when, e.g., crawling vocabularies to gather specific information.

>>>>

\subsubsection{Omitted Language Tags}
Some controlled vocabularies contain literals in natural language without information about what language has actually been used. Using private language tags (see RFC3066) or omitting them at all limits, e.g., language-dependent queries. To identfy omitted language tags we iterate over all conceptual resources in $C$ and collect those that have relations to untyped literals lacking language information as described above.

\subsubsection{Incomplete Language Coverage}
According to the previous criterion, vocabularies should combine plain literals with language tags. Furthermore, to support, e.g., internationalization or translation use cases, every concept should have defined literal values in the same set of languages than the other concepts. Calculation of the affected resources is done by iterating over all conceptual resources in $C$, building a global list of distinct languages throughout all literals assigned to every concept. Additionally, a list of used languages per conceptual resource is created. The criterion identifies all conceptual resources having a language list not equal to the global language list.

\subsubsection{Undocumented Concepts}
\cite{Aitchison2000} cites Svenonius to advocate ``inclusion of as much definition material as possible''. The SKOS language specification defines various properties that are intended to hold this kind of information, subsumed as ``documentation properties'' in the SKOS reference. To identify the affected conceptual resources, we iterate over all elements in $C$ to find those that are not related to one of the documentation properties.

\subsubsection{Potentially Semantically Related Concepts} \todo{All}{reformulate criterion title}
This criterion is an extension of a statement from the SKOS primer, recommending ``no two concepts have the same preferred lexical label in a given language when they belong to the same concept scheme''. Given the fact that (i) many vocabularies don’t make use of concept schemes and (ii) also identical altLabels or hiddenLabels can have a negative effect in some use-cases (e.g, auto-completion of concept labels based on user input), we follow a more general approach: identification of all pairs of concepts with their respective prefLabel, altLabel or hiddenLabel meeting a threshold of a certain similarity function $sim:LV \times LV \rightarrow [0,1]$. For the analysis of the criterion in this document, we define the similarity function  to check for case-insensitive string equality and the threshold to equal 1. Calculation of this criterion is done by applying the similarity function to every label of every possible pair of conceptual concepts in $C$ and filtering those that meet or exceed a specified threshold.

@@

\subsubsection{Loose Concepts} is a criterion motivated by the notion of ``orphan terms'' in the literature~\cite{Hedden2010}, which are are terms without any associative or hierarchical relationships. Checking for them is common in existing thesaurus development applications and also suggested by \cite{Z39.19:2005}. When we apply this to SKOS vocabularies, a loose concept is a concept that has no semantic relations with other conceptual resources and thus do not occur in any of the pairs yielded by $SR$. Although they might have attached lexical labels, they lack valuable context information that is considered essential for, e.g., search query expansion. Calculation can be done by iterating over all elements in $C$, filtering those not having a semantic relation to any other conceptual resource.

@@

\subsubsection{Omitted Top Concepts}
The SKOS language provides ConceptSchemes which are a facility for grouping related concepts. In order to provide entry points to such a group of concepts, one or more concepts can be marked as top concepts. This helps to provide ``efficient access'' (SKOS primer) and simplifies orientation in the vocabulary. To calculate this criterion, we identify all ConceptSchemes $I_{CEXT}(\texttt{skos:ConceptScheme}^\mathcal{I})$ not contained in any of the pairs in $I_{EXT}(\texttt{skos:hasTopConcept}^\mathcal{I})$.

@@

\subsubsection{Top Concept Having Broader Concepts}
\cite{Allemang2011} proposes to ``not indicate any concepts internal to the tree as top concepts'', i. e. top concepts should not have broader concepts. This criteria identifies all elements from $C$ that occur in both $I_{EXT}(\texttt{skos:topConceptOf}^\mathcal{I})$ and as first element of each pair from $I_{EXT}(\texttt{skos:broader}^\mathcal{I})$.

@@

\subsubsection{Missing In-Links}
A periodically updated version of a diagram visualizing ``the data sets in the LOD cloud as well as their interlinkage relationships'' is publised online\footnote{\url{http://www4.wiwiss.fu-berlin.de/lodcloud/state/}}. In these diagrams, dbpedia is shown as the most intensely linked resource, i.e. a large number of vocabularies link to resources in that dataset. Therefore a dataset with a large number of other datasets referencing it, can be assumed to be of great value for the community. Accordingly, estimating the number of In-Links of SKOS concepts defined in a vocabulary (henceforth called the in-degree of a concept) would give some hint about how established or accepted a controlled vocabulary is. A method to carry out these estimations would be utilizing existing Linked Data indices like, e.g., Sindice\footnote{\url{http://sindice.com/}} that provide a SPARQL endpoint. In order to estimate the number of In-Links for a conceptual resource, we iterate over all elements in $C$ and query the Sindice SPARQL endpoint for triples containing the concept's URI in the object part. Those concepts having an empty query result, are missing any In-Links and are thus included in the result set of this criterion.

\subsubsection{Missing Out-Links}
The Web of Data consists of globally interconnected datasets, ``enabling seamless connections between data sets''\cite{Heath2011}. These links can be established between various SKOS vocabularies by Out-Links, i.e., references from conceptual resources in the ``source vocabulary'' $V$ to resources identified by an HTTP URI and available at a different host. Similar to the criterion above, this criterion identifies the set of all conceptual resources that lack such Out-Links. Calculation of the result set is done by iterating over all elements $c \in C$ and investigating the host part of all related resources. If none of these host names differs from the host part of $c$, it has no Out-Links and is thus included into the resources identified by this criterion.

@@

\subsubsection{Link Target Unavailability}
This criterion identifies a set of all resources that have been identified as unavailable, i.e. resolving their URI leads to an erroneous HTTP response or no response at all. An erroneous HTTP response in that case can be defined as a response code (after a possible redirection) other than 200. Just as in the ``traditional'' Web, these ``broken links'' hinder information gathering and should be avoided. Calculation of this criterion is performed by iterating over all resources in $IR$, performing an HTTP request and include those resources in the result set that are unavailable.

@@

\subsubsection{Undefined SKOS Resource Usage}
Some vocabularies reference resources in the SKOS namespace that cannot be resolved. A common reason for this is, that there seems to exist a misunderstanding about the correct way of introducing new classes or properties in an OWL document. In some cases, vocabularies ``invent'' new terms in the SKOS namespace in order to meet certain requirements of the publishing organizations that cannot (yet) be expressed within the SKOS schema. Another reason for the presence of unresolvable resource references in the SKOS namespace is that the vocabulary might be outdated. With the SKOS W3C recommendation maturing, some properties have been removed from the current version of the schema. Vocabularies using these ``deprecated'' properties hence contain parts of their information in a non ``standard"-compliant way, having negative effects when, e.g., crawling vocabularies to gather specific information. A method to identify these undefined SKOS resources is to iterate over all resources in $IR$, including (i) those resources in the resulting set that are contained in the list of deprecated resources or (ii) have an URI in the SKOS namespace but are not defined in the SKOS ontology.
